---
title: "Reproducible Research: Peer Assessment 1"

---


## Pre-processing and clean-up

```{r, echo=TRUE}
##Required Libraries
library(ggplot2)
library(dplyr)

##Download File
zipurl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(zipurl, "temp.zip", mode="wb")
unzip("temp.zip", "activity.csv")
dd <- read.table("activity.csv", sep=",", header=T)

##Quick Look
summary(dd)

## subset data for complete cases only
d <- d[complete.cases(dd), ] ## drops missing values
```

## What is mean total number of steps taken per day?

```{r, echo=TRUE}
## summarize data
ds <- d %>% group_by(date) %>% summarise(sumsteps=sum(steps))

## histogram
dsum <- ggplot(data=ds, aes(x=sumsteps))
dsum + geom_histogram(aes(fill=..count..), binwidth=3000)

## mean number of steps taken per day
dmean <- mean(ds$sumsteps) ## mean of total number of steps
dmean

dmedian <- median(ds$sumsteps) ## median of total number of steps
dmedian
```

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis). Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r, echo=TRUE}

## summarize and plot
dint <- d %>% group_by(interval) %>% summarise(intmean=mean(steps))

dintline <- ggplot(data=dint, aes(x=interval, y=intmean))
  dintline + geom_line()

## max five minute interval
subset(dint, intmean == max(intmean), select=c(interval, intmean))
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs). Then create a new dataset that is equal to the original dataset but with the missing data filled in.

**Processing**
``` {r, echo=TRUE}
nas <- is.na(d) ## create NA index
dcln <- merge(d, dint, by="interval") ## merge in NA index
dcln$stepsx <- as.numeric(dcln$steps)
my.na <- is.na(dcln$steps)
dcln$stepsx[my.na] <- dcln$intmean[my.na] ## replace NAs with mean value
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Histogram of total number of steps taken per day

```{r, echo=TRUE}
dclns <- dcln %>% group_by(date) %>% summarise(sumstepsx=sum(stepsx)) ## summary

dclnshist <- ggplot(data=dclns, aes(x=sumstepsx))
dclnshist + geom_histogram(aes(fill=..count..), binwidth=3000)

dmeanx <- mean(dclns$sumstepsx) ## mean of total number of steps
dmeanx

dmedianx <- median(dclns$sumstepsx) ## median of total number of steps
dmedianx
```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

The differences are clear when looking at the two line plots below.

```{r, echo=TRUE}

## create weekeday variable and subset
dcln$week <- weekdays(as.Date(dcln$date)) ## create weekday / weekend variable

dclnweek <- subset(dcln, week == "Monday" | week == "Tuesday" | week == "Wednesday" |
                     week == "Thursday" | week == "Friday") ## week subset
dclnwend <- subset(dcln, week == "Saturday" | week == "Sunday") ## weekend subset

## summarize data
dclnweeksum <- dclnweek %>% group_by(interval) %>% summarize(sumstepsx=sum(stepsx))
dclnwendsum <- dclnwend %>% group_by(interval) %>% summarize(sumstepsx=sum(stepsx))

## differences between weekday and weekend activities
par(mfrow=c(2, 1))
plot(dclnweeksum$sumstepsx, type="l", col="blue", xlab="Interval", ylab="Number of Steps")
  title(main="Weekday", font.main=2)
plot(dclnwendsum$sumstepsx, type="l", col="blue", xlab="Interval", ylab="Number of Steps")
  title(main="Weekend", font.main=2)

```














